---
title: "NetCDF - Low Level usage in R"
author: Michael.Sumner@aad.gov.au
output: 
    ioslides_presentation:
    css: myslides.css
---

## R packages for NetCDF

- RNetCDF
- ncdf 
- ncdf4
- rgdal


```{r,echo=FALSE}
library(RNetCDF)
f <- "avhrr-only-v2_20140301.nc"
dp <- "~/ncdata"
```

## Typical session

```{r, eval=FALSE}
dp <- "~/ncdata"
f <- "avhrr-only-v2_20140301.nc"
library(RNetCDF)
nc <- open.nc(file.path(dp, f))
print.nc(nc)
## do other stuff
close.nc(nc)
```

```{r,echo=FALSE,eval=FALSE}
dp <- "~/ncdata"
f <- "avhrr-only-v2_20140301.nc"
library(RNetCDF)
nc <- open.nc(file.path(dp, f))
sink(sprintf("%s", gsub(".nc$", ".txt", f)))
print.nc(nc)
sink(NULL)
## do other stuff
close.nc(nc)
```

Source file has variables **sst** as well as **anom**, **err**, and **ice**. 

Daily sea surface temperature, Reynolds, et al.(2007) Daily High-resolution Blended Analyses. 

[avhrr-only-v2_20140301.ncdump](avhrr-only-v2_20140301.txt)



## Now what? 

Proceed with gay abandon. (Beware of large variables.)

```{r}
nc <- open.nc(file.path(dp, f))
lons <- var.get.nc(nc, "lon")
lats <- var.get.nc(nc, "lat")
dat <- var.get.nc(nc, "sst")
range(dat, na.rm = TRUE)
```

- It's rarely simple.
- Inspect output of ncdump.
- Check data ranges, orientation, missing values. 
- Code can interrogate **attributes** (the metadata) for sophisticated automation.

## Checks and visualization

Extremely important for checking assumptions/results. 

```{r}
m <- matrix(c(1, 2), 1, 2); layout(m)
plot(lons)
plot(lats)
```

## Visualization

Image of the raw matrix. 

```{r}
image(dat)
```



## Reading subsets

```{r}
xindex <- which(lons >= 130 & lons <= 190)
yindex <- which(lats >= -65 & lats <= -20)
lons2 <- var.get.nc(nc, "lon", start = xindex[1], count = length(xindex))
lats2 <- var.get.nc(nc, "lat", start = yindex[1], count = length(yindex))
```

```{r}
## WATCH OUT - degenerate dimensions (starts at 1, has length 1)
indexstart <- c(xindex[1], yindex[2], 1, 1)
indexcount <- c(length(xindex), length(yindex), 1, 1)
dat2 <- var.get.nc(nc, "sst", start = indexstart, count = indexcount)
```

## Visualization

```{r}
d <- list(x = lons2, y = lats2, z = dat2)
image(d)
```

## Bumps in the road

- multiple variables in one file, or across files

```{r}
## horizontal and vertical components of velocity
fu <- "uwnd.10m.gauss.2013.nc"
fv <- "vwnd.10m.gauss.2013.nc"

unc <- open.nc(file.path(dp, fu))
## should check the "v" file has the same stuff
vnc <- open.nc(file.path(dp, fv))
```

Daily Forecast of U[V]-wind at 10 m, NCEP/DOE AMIP-II Reanalysis (Reanalysis-2) Daily Averages. 

## Bumps in the road

- latitude top-down (was bottom-up for SST)
- longitude is Pacific-view (could be Atlantic [-180, 180]) 

```{r}
lons <- var.get.nc(unc, "lon"); lats <- var.get.nc(unc, "lat")
layout(matrix(c(1, 2), 1, 2)); plot(lons); plot(lats)
```


## Dimensions and subset

```{r}
timeind <- 10
## U, V with dimensions [longitude, latitude, time]
indexstart <- c(1, 1, 1, timeind)
indexcount <- c(length(lons), length(lats), 1, 1)
uwnd <- var.get.nc(unc, "uwnd", 
                   start = indexstart, count = indexcount)
vwnd <- var.get.nc(vnc, "vwnd", 
                   start =indexstart, count = indexcount)

```

## Scaling

```{r}
## uwnd:missing_value = 32766 ;
## uwnd:add_offset = 207.65 ;
## uwnd:scale_factor = 0.01 ;
   
## scaling from integer values (saves space, preserves precision)
uwnd <- uwnd * 0.01 + 207.65
vwnd <- vwnd * 0.01 + 207.65  ## best check they really are the same

##     time:units = "hours since 1800-1-1 00:00:0.0" ;
##     time:delta_t = "0000-00-01 00:00:00" ;
epoch <- ISOdatetime(1800, 1, 1, 0, 0, 0, tz = "GMT")
time <- var.get.nc(unc, "time") * 3600 + epoch
```

## Visualization 

Note top-down orientation: 

- reverse latitudes
- reverse order of columns. 
- 
```{r,echo=FALSE,message=FALSE}
library(sp);library(maptools)
```

```{r,eval=FALSE,echo=TRUE}
wspd <- sqrt(uwnd^2 + vwnd^2)[,ncol(uwnd):1]
image(lons, rev(lats), wspd)
title(format(time[timeind]))
library(maps);map("world2", add = TRUE)
```

## Visualization


```{r,eval=TRUE,echo=FALSE}
wspd <- sqrt(uwnd^2 + vwnd^2)[,ncol(uwnd):1]
image(lons, rev(lats), sqrt(uwnd^2 + vwnd^2)[,ncol(uwnd):1])
title(format(time[timeind]))
library(maps);map("world2", add = TRUE)
```

## Obstacles 

- Juggling file connections
- Discovery of names
- Internal conventions for orientation, organization
- Scaling to real values
- Handling missing values
- Extremely low-level interface



## UI differences in packages

Open, get variable, attribute inquiry, print: same things different names

- RNetCDF: open.nc, var.get.nc, att.get.nc, print.nc
- ncdf:   open.ncdf, get.var.ncdf, att.get.ncdf, print.ncdf
- ncdf4:  nc_open, ncvar_get, ncatt_get, print.ncdf4

(Also *write*, *copy*, *put*, *rename*, *delete*, etc.)

Interface for GDAL is pretty different. 

- rgdal:  GDAL.open, "[-methods", GDALinfo

## R packages for NetCDF

RNetCDF, ncdf, ncdf4, rgdal

Which is best? Depends: 

- how package/s built (source, CRAN binaries, distro binaries, etc.)
- inter-related with HDF4/5, OpenDAP, Thredds, compilers, etc. 
- important especially if you need to **write** NetCDF

No CRAN binary provides NetCDF-4. 

None read some NetCDF-4 models, such as HDF-like groups. 

We need a new-generation R NetCDF package, built via SWIG, Rcpp, ? 


## High-level in R

Read with full interpretation, crop. 

```{r,eval=FALSE}
library(raster)
## lonmin, lonmax, latmin, latmax
ex <- extent(130, 190, -65, -20)
sst <- crop(raster("avhrr-only-v2_20140301.nc"), ex, snap = "out")
## apply the same to "anom" variable
anom <- crop(raster("avhrr-only-v2_20140301.nc", varname = "anom"), 
             ex, snap = "out")
```

## Raster package

```{r}
library(raster)
r <- sqrt(raster(file.path(dp, fu))^2 + raster(file.path(dp, fv))^2)
plot(r, col = heat.colors(25, alpha = 0.75))
title("wind speed (m/s)")
```

## Taking it a step further 

Leveraging the *raster* package with interpretation of variables, levels, time steps, missing  values, scaling/offset, we add query by date and simple options against controlled collections. 

```{r,eval=FALSE}
library(raadtools)
sst <- readsst(time[timeind])
windmag <- readwind(time[timeind], lon180 = TRUE, magonly = TRUE)
##plot(sqrt(x[[1]]^2 + x[[2]]^2), col = heat.colors(25))
plot(windmag, col = heat.colors(25, alpha = 0.75))
contour(sst, add = TRUE, col = "white")
library(maptools)
data(wrld_simpl)
plot(wrld_simpl, add = TRUE)
```

 (shameless plug for in-development AAD/ACE project *raadtools*)



##  Wind magnitude with SST contours

```{r,eval=TRUE,echo=FALSE}
library(raadtools)
sst <- readsst(time[timeind])
windmag <- readwind(time[timeind], lon180 = TRUE, magonly = TRUE)
##plot(sqrt(x[[1]]^2 + x[[2]]^2), col = heat.colors(25))
plot(windmag, col = heat.colors(25, alpha = 0.75))
contour(sst, add = TRUE, col = "white")
library(maptools)
data(wrld_simpl)
plot(wrld_simpl, add = TRUE)
```


