---
title: "3 SQL use cases - GIS, NetCDF and dplyr"
author: "Michael Sumner"
date: "09/05/2015"
output: html_document
---

Structured Query Language (SQL) is extremely versatile and is available in a wide
range of software environments. Its main strengths are 

Here we build three use cases to illustrate the breadth of application. 

# GIS intersection queries

SQL is a fairly standard way of working with GIS data, also available in GDAL/OGR and QGIS, and databases with spatial extensions (SQL Server, PostGIS/PostgreSQL, Oracle, Spatialite, Access (PGDB)). 

The tasks is to classify interesting marine areas for mineral exploitation, we want to cross-tabulate areas of the marine environment by seafloor depth, arbitration zones, and habitat zones. 

Here we import a standard topography data set Etopo2 into Manifold as a surface, and apply its georferencing metadata. 

- rename to Etopo2
- set georeferencing (offsets slightly out because of NetCDF centre/corner ambiguity)
- untick "shading"

We like to work in a sensible map projection, so first crop to our extents in the Southern Ocean and reproject. 

We can select pixels and crop using the UI, but we choose SQL instead for the selection. 

```{sql,eval=FALST}
UPDATE [Etopo2] SET [Selection (I)] = True WHERE [Latitude (I)] <= -40; 
```

Use TransformWith script?


Create a map with the topography and choose a projection, once we are happy we can apply the projection to the data. 

Contour the bathymetry into a range of depth classes. 

-200
-500, 
-1000
-2000
-3000

Now we need our habitat and political layers. 

Walrus - draw some on, call them Walrus1-4 for eg.

CCAMLR - delete all columns bar asdlabel


Use Identity or Intersect?

Need to simplify the topo data first.

Finally we have a single query to cross tabulate our habitat and political zones by bathymetry class: 

```{sql,eval=FALSE}
TRANSFORM Sum([Area (I)])/1e6   SELECT [Walrus], [asdlabel] FROM [Subareas Drawing 3] GROUP BY [Walrus], [asdlabel]  PIVOT [Depth] 
```

# NetCDF data model

```{r}
library(raadtools)
cfiles <- subset(currentsfiles(), date > (Sys.time() - 365.25 * 24 * 3600))

i <- 1
library(RNetCDF)
nc <- open.nc(cfiles$fullname[i])
## lon, lat, time shared by u and v
library(dplyr)
dm <- rep(NA_integer_, 3L)
## u/v is lon, lat, time
lon <- var.get.nc(nc, "lon")
dm[1] <- length(lon)
lon <- data_frame(lon = lon, id = seq_along(lon))
lat <- var.get.nc(nc, "lat")
dm[2] <- length(lat)
lat <- data_frame(lat = lat, id = seq_along(lat))
time <- var.get.nc(nc, "time")
dm[3] <- length(time)
time <- data_frame(time = time, id = seq_along(time))
## tables are lon, lat, time, uv

uv <- data_frame(u = as.vector(var.get.nc(nc, "u")), v = as.vector(var.get.nc(nc, "v")), 
                 lon_id =  rep(lon$id, dm[2]), lat_id = rep(lat$id, each = dm[1]),  time_id = rep(i, prod(dm[1:2])))

test <- inner_join(inner_join(inner_join(uv, lon, by = c("lon_id" = "id")), lat, by = c("lat_id" = "id")), time, by = c("time_id" = "id"))
test1 <- test %>% filter(!is.na(u) & !is.na(v))
## compare
bu <- brick(cfiles$fullname[i], varname = "u")
bv <- brick(cfiles$fullname[i], varname = "v")
plot(sqrt(bu^2 + bv^2), col = topo.colors(25))
scl <- function(x) (x - min(x, na.rm = TRUE))/diff(range(x, na.rm = TRUE))
plot(test1$lon, test1$lat, pch = ".", col = topo.colors(25)[scl(sqrt(test1$u^2 + test1$v^2)) * 24 + 1])


```

```{r}
d <- vector("list", nrow(cfiles))
time <- vector("list", nrow(cfiles))
for (i in seq(nrow(cfiles))) {
  nc <- open.nc(cfiles$fullname[i])
  uv <- data_frame(u = as.vector(var.get.nc(nc, "u")), v = as.vector(var.get.nc(nc, "v")), 
                 lon_id =  rep(lon$id, dm[2]), lat_id = rep(lat$id, each = dm[1]),  time_id = rep(i, prod(dm[1:2])))
  tval <- var.get.nc(nc, "time")
  close.nc(nc)
  uv <- uv %>% filter(!is.na(u) & !is.na(v))
  d[[i]] <- uv
  time[[i]] <-  data_frame(time = tval, id = i)
  print(i)
}

d <- do.call(bind_rows, d)
time <- do.call(bind_rows, time)

test <- dw %>% filter(time_id == 1) %>%  inner_join(lon, by = c("lon_id" = "id")) %>% inner_join(lat, by = c("lat_id" = "id"))
plot(test$lon, test$lat, pch = ".", col = topo.colors(25)[scl(sqrt(test$u^2 + test$v^2)) * 24 + 1])

# dplyr
